---
title: ""
---

## Week 7

### Day 1

#### [Today]{style="color:red;"}

-   Student-led paper discussion

-   Lecture (Code): <a href="/classfiles/Bayesian.Hierarchical.Model.zip">Bayesian Hierarchical Model Files</a>

-   Install R packages:

```{r, eval=FALSE, echo=TRUE}
# CRAN packages
install.packages(c("brms","bayesplot"))
```

#### [Assigned]{style="color:red;"}
 -   Readings for next class: 
    - <a href="/readings/Gerber et al. 2014.pdf">Gerber et al. 2014</a>

 
- Background Reading (not required): 
  - [Kurz, 2023; Doing Bayesian Data Analysis in brms and the tidyverse](https://bookdown.org/content/3686/) 


### Day 2 (lab)

#### [Today]{style="color:red;"}

-   Lecture: Continue with: <a href="/FW680A4/bayesian.occ.qmd">Bayesian Occupancy Models</a>

-   Downloads: <a href="/classfiles/occlab/week7files.zip">Files</a>

-   Install R packages:

```{r, eval=FALSE, echo=TRUE}
# CRAN packages
install.packages(c(""))
```


#### [Assigned]{style="color:red;"}



<!-- -   Readings for next class:  -->
<!--     - <a href="https://statistics4ecologists-v2.netlify.app/mmi#mmi">Fieberg, Ch. 8</a> -->






