---
title: ""
---

## Week 9

### Day 1

#### [Today]{style="color:red;"}

-   Downloads: <a href="/classfiles/occlab/hbocclab.zip">Files</a>

-   Install R packages:

```{r, eval=FALSE, echo=TRUE}
# CRAN packages
install.packages(c("ubms","unmarked"))
```

#### [Assigned]{style="color:red;"}
-   Readings for next class:
    - <a href="/readings/Royle et al 2018.pdf">Royle et al. 2018</a>


- Background Reading (not required):
  - <a href="/readings/Borchers and Marques 2017.pdf">Borchers and Marques 2017</a>
  - <a href="/readings/Royle et al 2011.pdf">Royle et al 2011</a>

 
### Day 2 (lab)

#### [Today]{style="color:red;"}

##### **In Class Discussion of Royle et al. 2018**

- How are data different than a typical capture-recapture study?

- What is an 'activity center'?

- What makes this model 'spatial'? Think about this assumption. When will it be appropriate or not to assume?

- What are the two components that control 'detection probability'?

- If using a SCR model, how might your sampling strategy change?

- What is the exact meaning of abundance in this type of model?


-   Downloads: <a href="/classfiles/scrlab/scrlabfiles.zip">Files</a>


-   Install R packages:

```{r, eval=FALSE, echo=TRUE}
# CRAN packages
install.packages(c("secr","sf"))
```


#### [Assigned]{style="color:red;"}
- Readings for next class:
  - <a href="/readings/Zipkin and Saunders 2018.pdf">Zipkin and Saunders 2018 (lines 153-208, AKA section 3)</a>
  - <a href="/readings/Zipkin et al. 2021.pdf">Zipkin et al. 2021</a>
  - <a href="/readings/Powell and Gale 2015 Ch 18.pdf">Powell and Gale ch 18</a>


<!-- - Background Reading (not required): -->
  <!-- - <a href="/readings/Borchers and Marques 2017.pdf">Borchers and Marques 2017</a> -->






