---
title: ""
---

## Week 7

### Day 1

#### [Today]{style="color:red;"}

-   Please bring laptop today

-   Lecture (Code): <a href="/classfiles/hbayes/Bayesian.Hierarchical.Model.zip">Bayesian Hierarchical Model Files</a>

-   Install R packages:

```{r, eval=FALSE, echo=TRUE}
# CRAN packages
install.packages(c("brms","bayesplot"))
```

#### [Assigned]{style="color:red;"}
 -   Readings for next class: 
    - <a href="/readings/Kery and Schaub Ch13.pdf">Kery and Schaub Ch13</a>

 
### Day 2 (lab)

#### [Today]{style="color:red;"}

-   Discuss Project

-   Lecture: Board Work

-   Downloads: <a href="/classfiles/occlab/week7files.zip">Files</a>

-   Install R packages:

```{r, eval=FALSE, echo=TRUE}
# CRAN packages
install.packages(c("ubms","unmarked"))
```


#### [Assigned]{style="color:red;"}
 -   Readings for next class: 
 - <a href="/readings/">Assigned Reading on SDM</a>




<!-- -   Readings for next class:  -->
<!--     - <a href="https://statistics4ecologists-v2.netlify.app/mmi#mmi">Fieberg, Ch. 8</a> -->






