---
title: ""
---

## Week 7

### Day 1

#### [Today]{style="color:red;"}

-   Please bring your laptop today

-   Lecture (Code): <a href="/classfiles/hbayes/Bayesian.Hierarchical.Model.zip">Bayesian Hierarchical Model Files</a>

-   Install R packages:

```{r, eval=FALSE, echo=TRUE}
# CRAN packages
install.packages(c("brms","bayesplot"))
```

#### [Assigned]{style="color:red;"}
 -   Readings for next class: 
    - <a href="/readings/Kery and Schaub Ch13.pdf">Kery and Schaub Ch13 (sections 13.1 to 13.4)</a>

 
### Day 2 (lab)

#### [Today]{style="color:red;"}

-   Discuss Project

-   Lecture: Board Work

-   Downloads: <a href="/classfiles/occlab/week7files.zip">Files</a>

-   Install R packages:

```{r, eval=FALSE, echo=TRUE}
# CRAN packages
install.packages(c("ubms","unmarked"))
```


#### [Assigned]{style="color:red;"}

-   Readings for next class:
    -   <a href="/readings/Widmer et al 2021.pdf">Widmer et al 2021</a>
    -   <a href="/readings/Guillera‐Arroita et al. 2015.pdf">Guillera‐Arroita et al. 2015</a>

- Background Reading (not required):
  -   <a href="/readings/Soley-Guardia et al. 2024.pdf">Soley-Guardia et al. 2024</a>
  - <a href="/readings/Valavi et al. 2021.pdf"> Valavi et al. 2021 (assessment of modeling methods)</a>.



