---
title: "Matrices"
author: "Brian D. Gerber"
date: "2026-06-23"
output: 
  html_document:
    toc: true
---


<style type="text/css">
  body {
    font-size: 20px; /* Changes main body text */
  }
  code.r {
    font-size: 20px; /* Changes code chunk text */
  }
  pre {
    font-size: 20px;
  }
  h2 {
  color: #800080;
}
  h1 {
  color: #FFAE42;
}
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## The Setup

The goal of this script is to help understand matrix algebra and notation in linear and generalized linear models.

Imagine there are $n=10$ data points and you want to fit a linear regression model and predict $\hat{y}$ from the estimated coeffients ($\boldsymbol{\beta}$) and covariates with the design matrix ($\textbf{X}$).

We will ignore the estimation of coeffients and just pretend that we know what they are.

Lets setup the coeffients as one intercept and two slopes ($\boldsymbol{\beta}$)
```{r}
 betas = matrix(c(2,1,5),nrow=3)
 betas
```

Next, create a design matrix that has one intercept and two continuous covariates
```{r} 
 set.seed(34524254)
 X = matrix(cbind(rep(1,10),rnorm(10,2,1),rpois(10,10)),nrow=10)
 head(X)
```

If I want to multiple the coeffients by the covariates and intercept to predict $\hat{y}$, how do I relate $\textbf{X}$ and ($\boldsymbol{\beta}$)?


First, what are the dimensions of the matrices...

```{r}
 dim(X)
 dim(betas)
```

Can we multiply them together and what does that even mean?
```{r, error=TRUE}
 X*betas
 betas*X
```


No, we can't do that.

Rather, we need to do matrix multiplication (called a dot product ($\cdot$)), as

## Matrix multiplication

```{r}
 X%*%betas
```
Notice that the dimensions of the output is the number of samples, n = 10

But what is the dot product exactly doing?

Lets take one row at a time of ($\boldsymbol{\beta}$). Start with the first row

```{r}
 X[1,]
```
We want to multiply betas[1] by 1, and betas[2] by 2.91596 and betas[3] by 9
```{r}
  X[1,]*betas
```

And then we want to sum them...

```{r}
 sum(X[1,]*betas)
```

This is the same as the first element of 
```{r}
( X%*%betas)[1]
```

Remember what the $\boldsymbol{\beta}$ are. They are the partial effects of a known variable or covariate. The coeffients in columns 2 and 3 are slopes of a line or magnitude of effect that all together sum (linear approximation) to relate to the data, $y$. 


Go ahead and calculate manually the matrix multiplication of row 7 of $\textbf{X}$ and the $\boldsymbol{\beta}$

Is your answer the same as (X%*%betas)[7]?


Summary: So X%*%betas is multiplying all the elements of coeffients by each row of $\textbf{X}$ and then summing it. 

## Matrix Transpose

Transpose is an important matrix operation. It takes a matrix and turns it on its side.

```{r}
 X
 t(X)
```

Note the dimension change

```{r}
 dim(X)
 dim(t(X))
```

What about betas?

```{r}
 dim(betas)
 dim(t(betas))
```


If you multiply two matrices together, you should know the dimensions of the output matrix before the multiplication.

```{r}
X%*%t(X)
```

$\textbf{X}$ by $\textbf{X}^{t}$ is matrix multiplying a 10x3  matrix by a 3x10 matrix.

We know we can multiply them because the columns of the first matrix (3) is the same number as the rows of the second matrix (3). The matrix that is produced has the dimensions of rows of the first matrix (10) and the number of columns of the second matrix (10). Thus, the outcome is a 10x10 matrix

```{r}
 dim(X)
 dim(t(X))
 dim(X%*%t(X))
```


We can not multiple matrices that don't align in terms of the dimensions
```{r, error=TRUE} 
 X%*%X
```
Why is X%*%X non-conformable?

Can we always multiply a square matrix by itself? Try it.

## Identity Matrix

The Identity matrix is a special matrix, in which the diagonals are all one, and all other elements (off diagonals) are zeros.

```{r}
 diag(10)
 diag(3)
```

What is the outcome of multiplying $\boldsymbol{\beta}$ by an identity matrix
```{r} 
 diag(3)%*%(betas)
```

Nothing changed!

In GLM's an identity link is the identity matrix- basically, no transformation. In contrast to a log-transformation for the Poisson or the logit for the Binomial.

## Questions

If I multiply a matrix that is 8x10 times a matrix that is 8x10, what is the dimension of the produced matrix? See if you can figure it out.

If I multiply a matrix that is 9x3 times a matrix that is 3x1000, what is  the dimension of the produced matrix? See if you can figure it out.

Can we multiply a matrix by a scalar?
 
```{r}
 scalar=3
 diag(3)*scalar
 
# OR, how about
 X*100
```

Yes, we can do that, no problem.

Can we divide, add, and substract by a scalar?
 
```{r} 
 diag(3)/scalar

#OR
 diag(3)+1000
 diag(3)-1000
```

Yes, not a problem

Can we do these in a different order?

```{r}
 3/diag(3)
 1000+diag(3)
 1000-diag(3)
```

 Yes, but the order of operations matters, so the output is different. We also shouldn't divide by zero!

## Matrix operations
 
Does it matter what order of operations we do our matrix multiplication?
 
```{r} 
 A=matrix(c(10,2,1,5),nrow=2,ncol=2)
 B=matrix(c(1,5,2,10),nrow=2,ncol=2)
 
 A
 B

 A%*%B
 B%*%A
```

Are the output matrices the same? No!

We can't assume $A\cdot B$ is equal to $B\cdot A$ when dealing with matrices, which we can assume with regular multiplication: $3\times2 = 2\times3$

## An applied example

What if we don't have continiuos variables, but categorical?

Imagine a categorical variable (habtat) with three levels: grassland, forest, shrub

We want to relate elk density ($y$; our data) to habitat with these three levels. We thus need a  design matrix ($\textbf{X}$) that has 3 columns, to translate which row of data correspond to each level.

Lets consider 12 elk densities in which 4 of each correspond to each of the habitat levels

Here is the variable
```{r}
 habitat=factor(c(rep("Forest",4),rep("Grass",4),rep("Shrub",4)))
 habitat
```

Let's turn this into a design matrix
```{r}
 X=model.matrix(~habitat)
 X
```

This transformation is a really really important part of understanding all linear models. 

Notice that the first column is the intercept, the second and third columns are called partial intercepts or dummy codes to indicate the different levels.

For each row, there is a unique combination of 1's and 0's that indicate which is forest, grassland, and shrub. The first row is 1 0 0, and we know looking at 'habitat' that the first element is 'forest'. The 5th row is 1 1 0, which from 'habitat', we know is grassland. Lastly, row 9 is 1 0 1, which we can see is Shrub.

How do we interpret the intercept?

The intercept alone ( 1 0 0 ), indicates Forest because no matter what the other coefficient values are we will always only be adding 1*beta1 + 0*beta2 + 0*beta3, which is equal to just 1*beta1. Thus, the intercept and beta1 indicate the  density of elk in forest. 

What does the 2nd column indicate? The beta2 corresponds to the difference in elk density between forest and grassland, so if grassland has less elk than forests beta2 will be negative, and if grassland has more elk than forests it will be positive. If there is no difference, we would expect beta2 to be zero.

The same is true for column three. beta3 is the difference from the intercept (forest) because anytime we multiply the betas by the rows of X, we only ever add the  intercept and one of the other habitat types.

# What if you didn't care that much about forest, but wanted to know the  effect difference between grassland and shrub. Then, just use a different design matrix

# Now, lets we reassign which is the "reference" category 
 
```{r} 
 habitat2=relevel(habitat, ref="Shrub")
 X2=model.matrix(~habitat2)
 X2
```

Notice how the design matrix has changed. Now, we have the 1 0 0 be the last part of the data, thus the intercept corresponds to "Shrub", 

Let use a set of betas and get the prediction of y (elk density) using the new design matrix
 
```{r} 
 betas2=c(0.5,1,2)
 X2%*%betas2
```
 
By changing the design matrix ( ($\textbf{X}$)) you can manipulate what betas are estimate and what they mean, but not necessarily the predictions of y-hat. Below is an example.

## Design matrix and predictions

#This is our data...
```{r}
 y=c(3, 3, 3, 3, 1.5, 1.5, 1.5, 1.5, 1,  1,  1,  1)
```

Let's relate our data to two different betas and sets of design matrices  

```{r}
 y.hat.1=t(X%*%c(3,-1.5,-2))
 y.hat.2=t(X2%*%c(1,2,0.5))

 y==y.hat.1
 y==y.hat.2
 y.hat.1==y.hat.2
```
Is it an issue that our betas are not unique solutions, but can change depending on our design matrix?

No, not at all. We just need to be certain that we understand the design matrix so that we can understand the meaning of out betas. But, irregardless of their meaning, we still predict the same y-hat's.

When using a design matrix  ($\textbf{X}$) what is the definition of the intercept (beta1)?
This indicates the elk desnity in Forests

When using design matrix  ($\textbf{X2}$) what is the definition of the intercept (beta1)?
This indicates the elk desnity in Shrubs

When using design matrix  ($\textbf{X}$) what is the definition of slope 1 (beta2)?
It is the difference in elk density between forests and .....

When using design matrix  ($\textbf{X2}$) what is the definition of slope 1 (beta2)?
It is the difference in elk density between shrubs and .....

When using design matrix  ($\textbf{X}$) what is the definition of slope 2 (beta3)?

When using design matrix  ($\textbf{X2}$) what is the definition of slope 2 (beta3)?


## lm and glm

How does this all relate to linear modeling and generalized linear modeling?

Does using variable habitat or habitat2 change the inference on the coefficients?
Does the variable habitat or habitat2 change the predictions from the model?


Lets fit a linear regression model with the variable habitat
```{r}
 fit1=lm(y~habitat)
 fit1
 fit1$model
```

Do the predictions match y? 

```{r}
 predict(fit1)
 y
``` 
 
#Now use habitat2 instead of habitat. Did the coefficients change? What about the predictions?

```{r}
 fit2=lm(y~habitat2)
 fit2
 predict(fit2)
 y
```


Instead of doing y~habitat, lets specify the design matrix we want to use, X1 and X2.

If we fit the y data with the design matrix, do they match with above? Meaning do we get the same results by using habitat or by using X?
#Do we also get the same results by using habitat2 or by using X2?
```{r}
 lm(y~X+0)
 lm(y~X2+0)
```

Do the same as above but for glm..

```{r}
 fit3=glm(y~habitat,family=gaussian)
 fit3
 predict(fit3)
 y
```

Same results as y~habitat?

```{r}
 glm(y~X+0,family=gaussian)

 fit4=glm(y~habitat2,family=gaussian)
 fit4
 predict(fit4)
 y
```

Same results as y~habitat2?

```{r}
 glm(y~X2+0,family=gaussian)
```

## THE END