---
title: ""
format:
  revealjs:
    chalkboard: true
---

## NRS 520 {.smaller .scrollable}
```{=html}

<title> NRS 520 Course Paage</title>

<style type="text/css">

body, td {
   font-size: 14px;
}
</style>

```

::: {.panel-tabset}

### Start 
<p><a href="/Schedule/Schedule.html">Schedule/Syllabus</a></p>

Install the most versions of <a href="https://cran.r-project.org/">R</a> and <a href="https://www.rstudio.com/products/rstudio/download/">RStudio</a>.

<a href="https://rstudio.uri.edu/">CELS RStudio Server</a>

<a href="https://github.com/">Github</a>

<br>

#### Notes

-   To download lectures. Go to lecture html. Click 'e'. This should connect all slides into one page. Right Click Mouse on page. Click 'Print". Choose printer "Adobe PDF"; might need Adobe Acrobat installed. 

### Learning R

<a href="http://preludeinr.com/r-basics/">Prelude In R</a>

<a href="https://datacarpentry.org/R-ecology-lesson/">Data Carpentry</a>

<a href="https://intro2r.com/">Intro to R</a>

<a href="https://paulvanderlaken.com/2017/08/10/r-resources-cheatsheets-tutorials-books/">Many Resources</a>

<a href="https://ourcodingclub.github.io/tutorials/intro-to-r/">Our Coding Club</a>

<a href="http://swcarpentry.github.io/r-novice-inflammation/">SW Carpentry</a>

<a href="https://www.jeffdoser.com/files/for875/_book/">Jeff Doser Book</a>


### Week 1 

-   Lecture: <a href="/Week 1/Introduction.html">1) Introduction</a>, <a href="/Week 1/Information.html"> 2) Information</a>

-   Lecture Code: <a href="/Week 1/Information.R">Information.R</a>

-   Markdown Example: <a href="/Week 1/Markdown and R Stuff.Rmd">1) Markdown Example File</a>, <a href="/Week 1/Markdown-and-R-Stuff.html">2) Compiled HTML</a>

-   Assignment: <a href="/Week 1/Assignment.Rmd">1) Markdown File</a> <a href="/Week 1/Assignment.html">2) Compiled HTML</a>

-    <a href="/Week 1/getting started with statistics.pdf">Assigned Reading</a>

### Week 2

We will be working in R and Markdown. For class, we will be going through (at least) modules 1 and 2. We will also do a bit on Markdown.

-   R Lecture Modules: <a href="/Week 2/R Class.zip">R Class.zip</a>

- R project organization example: <a href="/Week 2/Project Structure.zip">Project Structure.zip</a>

### Week 3
-   Install R packages:
```{r, eval=FALSE, echo=TRUE}
install.packages(c("pwr","ggplot2", "spsurvey", "foreach",
                   "sf", "doParallel", "grDevices","tictoc","plotly"))
```

-   Lecture: <a href="/Week 3/StudyDesign.html">Study Design</a>

-   Lecture Code: <a href="/Week 3/StudyDesign.R">StudyDesign.R</a>


-   Download: <a href="/Week 3/RI Deer Sampling.zip">Zip</a>

-   Assignment (option 1): <a href="/Week 3/Assignment.design.Rmd">1) Markdown File</a> <a href="/Week 3/Assignment.design.html">1) Compiled HTML</a>

-   Assignment (option 2): <a href="/Week 3/Assignment.design2.Rmd">1) Markdown File</a> <a href="/Week 3/Assignment.design2.html">1) Compiled HTML</a>


-   <a href="/Week 3/Betts et al.pdf">Assigned Reading</a>

### Week 4

-   Install R packages:
```{r, eval=FALSE, echo=TRUE}
install.packages(c("visualize","plotly"))
```

-   Lecture: <a href="/Week 4/Probability.html">Probabilty</a>

-   Lecture Code: <a href="/Week 4/Probability.R">Probabilty.R</a>

-   Download: 

-   Assignment: <a href="/Week 4/Assignment.pdf.Rmd">1) Markdown File</a> <a href="/Week 4/Assignment.pdf.html">1) Compiled HTML</a>


-   <a href="https://medium.com/@srowen/common-probability-distributions-347e6b945ce4">Assigned Reading</a>

-   <a href="/Week 4/SuggestedReadings.zip">Suggested Readings</a>

### Week 5

-   Lecture: <a href="Week 5/linear.model.html">GLM</a>

-   Lecture Code: <a href="/Week 5/linear.model.R">linear.model.R</a>

-   Download: <a href="">NEED?</a>

-   <a href="/Week 5/Assignment.GLM.html">Assignment</a>

-   <a href="/Week 5/Ecology - 2021 - Tredennick.pdf">Assigned Reading</a>

-   <a href="/Week 5/Chapter 3. Linear Regression.pdf">Suggested Reading</a>


### Week 5
### Week 6
### Week 7
### Week 8
### Week 9
### Week 10
### Week 11
### Week 12
### Week 13
### Week 14

:::