---
title: ""
format:
  revealjs:
    chalkboard: true
---

## NRS 520 {.smaller .scrollable}
```{=html}

<title> NRS 520 Course Paage</title>

<style type="text/css">

body, td {
   font-size: 14px;
}
</style>

```

::: {.panel-tabset}

### Start 
<p><a href="/Schedule/Schedule.html">Schedule/Syllabus</a></p>

Install the most versions of <a href="https://cran.r-project.org/">R</a> and <a href="https://www.rstudio.com/products/rstudio/download/">RStudio</a>.

<a href="https://rstudio.uri.edu/">CELS RStudio Server</a>

<a href="https://github.com/">Github</a>

<br>

#### Notes

-   To download lectures. Go to lecture html. Click 'e'. This should connect all slides into one page. Right Click Mouse on page. Click 'Print". Choose printer "Adobe PDF"; might need Adobe Acrobat installed. 

### Project

-   Create a group of 3

-   The group should discuss ideas in quantitative ecology that you would all like to develop a class dedicated to. If you need help, schedule a time to talk with me.

-   Outline a plan: 
    -   Identify a reading that will be assigned prior to class along with relevant topic questions
    -   Identify the scope and depth of the lecture material (bullet points of what will be covered), 
    -   Identify what the goal is to be accomplished in terms of coding (bullet points) 
    - Decide on a general idea for a homework assignment. 

-  Once outlined, email me the plan with all group members names. I will look over the general plan and provide comments. 

-   From here, I expect you to move forward with your plan and reach out when you have questions. Arrange at least one time to walk me through the lesson plan prior to the class, such that there is enough time to implement any feedback.  


### Learning R

<a href="http://preludeinr.com/r-basics/">Prelude In R</a>

<a href="https://datacarpentry.org/R-ecology-lesson/">Data Carpentry</a>

<a href="https://intro2r.com/">Intro to R</a>

<a href="https://paulvanderlaken.com/2017/08/10/r-resources-cheatsheets-tutorials-books/">Many Resources</a>

<a href="https://ourcodingclub.github.io/tutorials/intro-to-r/">Our Coding Club</a>

<a href="http://swcarpentry.github.io/r-novice-inflammation/">SW Carpentry</a>

<a href="https://www.jeffdoser.com/files/for875/_book/">Jeff Doser Book</a>


### Week 1 

-   Lecture: <a href="/Week 1/Introduction.html">1) Introduction</a>, <a href="/Week 1/Information.html"> 2) Information</a>

-   Lecture Code: <a href="/Week 1/Information.R">Information.R</a>

-   Markdown Example: <a href="/Week 1/Markdown and R Stuff.Rmd">1) Markdown Example File</a>, <a href="/Week 1/Markdown-and-R-Stuff.html">2) Compiled HTML</a>

-   Assignment: <a href="/Week 1/Assignment.Rmd">1) Markdown File</a> <a href="/Week 1/Assignment.html">2) Compiled HTML</a>

-    <a href="/Week 1/getting started with statistics.pdf">Assigned Reading</a>

### Week 2

We will be working in R and Markdown. For class, we will be going through (at least) modules 1 and 2. We will also do a bit on Markdown.

-   R Lecture Modules: <a href="/Week 2/R Class.zip">R Class.zip</a>

- R project organization example: <a href="/Week 2/Project Structure.zip">Project Structure.zip</a>

-    <a href="/Week 2/A comparison of design‐based and model‐based approaches for finite population spatial.pdf">Assigned Reading</a>

### Week 3
-   Install R packages:
```{r, eval=FALSE, echo=TRUE}
install.packages(c("pwr","ggplot2", "spsurvey", "foreach",
                   "sf", "doParallel", "grDevices","tictoc","plotly"))
```

-   Lecture: <a href="/Week 3/StudyDesign.html">Study Design</a>

-   Lecture Code: <a href="/Week 3/StudyDesign.R">StudyDesign.R</a>


-   Download: <a href="/Week 3/RI Deer Sampling.zip">Zip</a>

-   Assignment (option 1): <a href="/Week 3/Assignment.design.Rmd">1) Markdown File</a> <a href="/Week 3/Assignment.design.html">1) Compiled HTML</a>

-   Assignment (option 2): <a href="/Week 3/Assignment.design2.Rmd">1) Markdown File</a> <a href="/Week 3/Assignment.design2.html">1) Compiled HTML</a>

-   Assignment (option 3): <a href="/Week 3/Assignment.design3.Rmd">1) Markdown File</a> <a href="/Week 3/Assignment.design3.html">1) Compiled HTML</a>

-   Assigned Readings <a href="https://medium.com/@srowen/common-probability-distributions-347e6b945ce4">1) Probability Distributions</a>, <a href="/Week 3/HobbsHooten_Ch_3.pdf"> 2) Probability First Principles</a>

-   <a href="/Week 3/SuggestedReadings.zip">Suggested Readings</a>

### Week 4

-   Install R packages:
```{r, eval=FALSE, echo=TRUE}
install.packages(c("visualize","plotly"))
```

-   Lecture: <a href="/Week 4/Probability.html">Probabilty</a>

-   Lecture Code: <a href="/Week 4/Probability.R">Probabilty.R</a>

-   Download: 

-   Assignment: <a href="/Week 4/Assignment.pdf.Rmd">1) Markdown File</a> <a href="/Week 4/Assignment.pdf.html">1) Compiled HTML</a>


-   Assigned Reading <a href="/Week 4/Betts et al.pdf">Betts et al. 2021</a> and <a href="/Week 4/Chapter 3. Linear Regression.pdf">Linear Regression Book Chapter</a> or <a href="https://mlu-explain.github.io/linear-regression/">Online Linear Regression</a>

### Week 5

No Class / Holiday

### Week 6

-   Install R packages:
```{r, eval=FALSE, echo=TRUE}
install.packages(c("sjPlot","ggplot2", "ggResidpanel"))
```

-   Lecture: <a href="/Week 5/regression.html">Regression</a>

-   Lecture Code: <a href="/Week 5/regression.R">regression.R</a>

-   Download: <a href="/Week 5/elephant.study.csv">csv</a>

-   Assignment: <a href="/Week 5/Assignment.Regression.Rmd">1) Markdown File</a> <a href="/Week 5/Assignment.Regression.html">1) Compiled HTML</a>

-   Assigned Readings: <a href="/Week 5/Ecology - 2021 - Tredennick.pdf">Assigned Reading</a> 



### Week 7

-   Install R packages:
```{r, eval=FALSE, echo=TRUE}
install.packages(c("equatiomatic","faux"))
```

-   Lecture: <a href="/Week 6/linear.model.html">GLM</a>

-   Lecture Code: <a href="/Week 6/linear.model.R">linear.model.R</a>

-   Assignment: <a href="/Week 6/Assignment.GLM.Rmd">1) Markdown File</a> <a href="/Week 6/Assignment.GLM.html">1) Compiled HTML</a>

-   Assigned Readings: <a href="/Week 6/Bayesian statistics and modelling.pdf">Assigned Reading</a> 

### Week 8

- Download/Install JAGS software [here](https://sourceforge.net/projects/mcmc-jags/). The version you pick will depend on your version of R. 


-   Install R packages:
```{r, eval=FALSE, echo=TRUE}
install.packages(c("rjags","bayesplot","rstanarm"))
```

-   Lecture: <a href="/Week 7/Bayesian.html">Bayesian</a>

-   Lecture Code: <a href="/Week 7/bayesian.r">Bayesian.R</a>

-   Data Download: <a href="/Week 7/counts.trees1.csv">Data Set 1</a>, and <a href="/Week 7/counts.trees2.csv">Data Set 2</a>

-   Jags Code Example: <a href="/Week 7/jags.model.script.r">Script</a>, and <a href="/Week 7/model.jags.r">Model</a>

-   Stan Code Example:  <a href="/Week 7/stan.code.r">Script</a>

-   Extra Model Files: MCMC Algorithim Example: <a href="/Week 7/binom.logit.script.R">R Script</a>, and <a href="/Week 7/binom.logit.mcmc.R">MCMC Algorithim Function</a>


-   Assignment: <a href="/Week 7/Assignment.Bayesian.Rmd">1) Markdown File</a> <a href="/Week 7/Assignment.Bayesian.html">1) Compiled HTML</a>

-   Re-read the previously assigned Reading: <a href="/Week 6/Bayesian statistics and modelling.pdf">Assigned Reading</a> 

- Find a scientific paper that uses Bayesian modeling/inference in your subject area of interest. Be prepared to talk about what the study was about, how the authors used priors, what they made inference on, and how clear were the elements of modeling inference/modeling/computation. 

### Week 9

<span style="color:red">No Class or New Homework</span>

### Week 10

-   Install R packages:
```{r, eval=FALSE, echo=TRUE}
install.packages(c("rjags","bayesplot","runjags","coda"))
```

-   Lecture: <a href="/Week 8/RandomEffect.html">Random Effect</a>

-   Lecture Code: <a href="/Week 8/random.effect.R">random.effect.R</a>


### Week 11

-   Install R packages:
```{r, eval=FALSE, echo=TRUE}
install.packages(c("rjags","bayesplot","runjags","coda"))
```


-   Data Simulation Code (developed previous class): <a href="/Week 9/Inclass.data.simulation.r">Inclass.data.simulation.r</a>

-   JAGS Models: <a href="/Week 9/model.JAGS.extra.noise.r">model.JAGS.extra.noise.r</a> and <a href="/Week 9/model.JAGS.random.slope.r">model.JAGS.random.slope.r</a>

-   Model Scripts: <a href="/Week 9/Model.Script.Extra.Noise.r">Model.Script.Extra.Noise.r</a> and <a href="/Week 9/Model.Script.Random.Slope.r">Model.Script.Random.Slope.r</a>

-   Assignment: <a href="/Week 9/Assignment.RandomEffect.Rmd">1) Markdown File</a> <a href="/Week 9/Assignment.RandomEffect.html">1) Compiled HTML</a>

### Week 12
Laken, Maegan, Owen, and John lecture/lab 

-   Install R packages:
```{r, eval=FALSE, echo=TRUE}
install.packages(c("unmarked"))
```


-   Assigned Readings: <a href="/Week 10/Efford 2012.pdf">Assigned Reading</a> and associated <a href="/Week 10/Questions.docx">questions</a>.



### Week 13

Tori, Liam, and Rebeca lecture/lab  

### Week 14
```{r, eval=FALSE, echo=TRUE}
install.packages(c("Hmisc","foreign","reshape2"))
```


Caroline, Corey, and Rea lecture/lab

:::