---
title: "Chapter 3 Problem"
author: "Brian Gerber"
date: "2025-08-01"
output: html_document
---


  
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Chapter 3, Problem 1

<style type="text/css">
  body{
  font-size: 14pt;
}
pre.r { /* Targets R code blocks */
      font-size: 12pt;
    }
code { /* Targets inline code */
      font-size: 12pt;
    }

</style>


From the data of parts (a) and (b) of Exercise 1 of Chapter 2, construct approximate 95% confidence intervals for the population total. Also construct
approximate 95% confidence intervals for the population mean per unit. On what assumptions or results is the confidence interval procedure based, and
how well does the method apply here?

### Part a

```{r}

# Data

y = c(0, 1, 1, 1, 1, 3, 1, 0, 4, 0)
N = 100
n = length(y)

#Estimated total population size
  tau = N*mean(y)

# Create 95% CIs for the population total
  
# Upper  
  tau + qt(0.975,df=n-1)*sqrt(N*(N-n)*(var(y)/n))

# Lower
  tau + qt(0.025,df=n-1)*sqrt(N*(N-n)*(var(y)/n))


# Create 95% CIs for the population mean

# Upper limit
  mean(y) + qt(0.975,df=n-1) * sqrt(((N-n)/N)*(var(y)/n))

# Lower Limit
  mean(y) + qt(0.025,df=n-1) * sqrt(((N-n)/N)*(var(y)/n))

```

### Part b


```{r}

# Data
  y <- c(1, 0, 1, 1, 3, 1, 1, 0, 2, 2)
  N = 100
  n = length(y)

#Estimated total population size
  tau = N*mean(y)

# Create 95% CIs for the population total
  
# Upper  
  tau + qt(0.975,df=n-1)*sqrt(N*(N-n)*(var(y)/n))

# Lower
  tau + qt(0.025,df=n-1)*sqrt(N*(N-n)*(var(y)/n))


# Create 95% CIs for the population mean

# Upper limit
  mean(y) + qt(0.975,df=n-1) * sqrt(((N-n)/N)*(var(y)/n))

# Lower Limit
  mean(y) + qt(0.025,df=n-1) * sqrt(((N-n)/N)*(var(y)/n))

```

