---
title: "Coding and Mathematical Definitions"
author: "Brian D. Gerber"
date: "December 10, 2024"
output: pdf_document
header-includes:
   - \usepackage{calc,amsmath,tikz}
   - \usetikzlibrary{matrix,calc}
---

```{r setup, include=FALSE}
options(tinytex.verbose = TRUE)
```

\vspace{12pt}
\begin{center}\textbf{\LARGE Definitions of R Code} \end{center}
\vspace{12pt}

\textbf{\large R script}

the lines of code and comments that you are writing (filename.R).  
\vspace{12pt}

\textbf{\large R project}

your R script, any variables you have created, and your current R environment (filename.Rproj).  
\vspace{12pt}

\textbf{\large R package}

a set of functions/code that you can load into your script (examples - dplyr, sp).  
\vspace{12pt}

\textbf{\large R Comment}: does not run as code, starts with #. You use comments to explain what your code is doing in plain langauge.  
\vspace{12pt}

\textbf{\large Object or Variable}

a unit of information that is stored in the workspace (computer memory) and can be recalled or manipulated. 'a' is an object. Specifically, a vector of length 3.  

```{r include=TRUE, tidy=TRUE,results='asis'}
a = c(3,2,1)
```
\ \
\textbf{\large Element}

a piece of information within an object. The 1st element of a is...

```{r include=TRUE, tidy=TRUE, eval=TRUE,results='asis'}
a[1]
```
\ \
\vspace{12pt}

\textbf{\large Function}

a command to take inputs (objects or elements) and manipulates it to provide an output, which can be saved as a new object.
```{r include=TRUE, tidy=TRUE, eval=TRUE,results='asis'}
fun.text=function(x){paste("Your input variable is ",x,sep="")}

fun.text(4)
fun.text("INPUT")
```
\ \
\vspace{12pt}

\textbf{\large Argument}

a specific command within a function. Many functional arguments are preset and do not need to be explitly stated.

```{r include=TRUE, tidy=TRUE, eval=TRUE,results='asis'}
x=c(1,2,3,NA)
mean(x)

#na.rm is an argument of the function mean to ignore the missing value. Its preset is FALSE. If you
#have na's you need to change the argument to 'TRUE'.
mean(x, na.rm=TRUE)

```

\vspace{12pt}



\begin{center}\textbf{\LARGE Types of R Objects and Mathematical Notation} \end{center}
\vspace{12pt}

\textbf{\large Vector}

1 row, many columns OR 1 column many rows. Can be numbers of characters.

```{r include=TRUE, tidy=TRUE, eval=TRUE,results='asis'}
v=c(4,1,3)
is.vector(v)
length(v)
```

Math notation (capitalized & lower case):
$\vec{v}$ \ or \ $\textbf{v}$

\ \

\begin{align}
  \textbf{v} &= \begin{bmatrix}
           4 &
           1 &
           3
\end{bmatrix}
\end{align}

\vspace{12pt}

\textbf{\large Matrix}

Generalization of vectors. Can have 1 or more rows and columns. Only numbers.

```{r include=TRUE, tidy=TRUE,eval=TRUE}

M=matrix(1:10,nrow=2)
M
dim(M)
is.vector(M)
is.matrix(M)
```

Math notation (capitalized & upper case):
$\textbf{M}$

\ \

\begin{align}
  \textbf{M} &= \begin{bmatrix}
           1 & 3 & 5 & 7 & 9 \\
           2 & 4 & 6 & 8 & 10
\end{bmatrix}
\end{align}

\ \

\vspace{12pt}

\textbf{\large Array}

Generalization of matrices. Can be n dimensional. Only numbers.

```{r include=TRUE, tidy=TRUE,eval=TRUE}

arr=array(1:100,dim=c(3,4,3))
arr
dim(arr)
is.vector(arr)
is.matrix(arr)
is.array(arr)
```

\begin{tikzpicture}[every node/.style={anchor=north east,fill=white,minimum width=1.4cm,minimum height=7mm}]
\matrix (mA) [draw,matrix of math nodes]
{
25 & 28 & 31 & 34 \\
26 & 29 & 32 & 35 \\
27 & 30 & 33 & 36 \\
};

\matrix (mB) [draw,matrix of math nodes] at ($(mA.south west)+(1.5,0.7)$)
{
13 & 16 & 19 & 22 \\
14 & 17 & 20 & 23 \\
15 & 18 & 21 & 24 \\
 \\
};

\matrix (mC) [draw,matrix of math nodes] at ($(mB.south west)+(1.5,0.7)$)
{
1 & 4 & 7 & 10 \\
2 & 5 & 8 & 11 \\
3 & 6 & 9 & 12 \\
};

\draw[dashed](mA.north east)--(mC.north east);
\draw[dashed](mA.north west)--(mC.north west);
\draw[dashed](mA.south east)--(mC.south east);
\end{tikzpicture}


\ \

\vspace{12pt}

\textbf{\large List}

Can store any type of objects together.

```{r include=TRUE, tidy=TRUE,eval=TRUE}
list1=vector("list",2)
list1[[1]]=v
list1[[2]]=M
list1[[3]]=arr

is.list(list1)
length(list1)
list1[[1]]
```

\vspace{12pt}

\begin{center}\textbf{\LARGE Important Functions} \end{center}

\textbf{\large For Loop} 

To do a task many times. Simple, but inefficient

```{r include=TRUE, tidy=FALSE,eval=TRUE}
#First, create a large matrix
x <- matrix(rnorm(400*4000), ncol=400)

#Second, create a vector to store results
mx <- rep(NA, nrow(x))

#Third, for each row of the matrix, find the maximum value and store it in mx using 
#a loop. We will do this from index 1 to the max number of rows (nrow(x)). We will 
#iterate using index i.

for(i in 1:nrow(x)){
  mx[i] <- max(x[i,])
}

```

\vspace{12pt}

\textbf{\large ifelse}

used to return an output based on specified conditions

```{r echo=TRUE, eval=FALSE,include=TRUE, tidy=FALSE}
# Syntax:
If (test_expression) {
  Statement 
}

```

```{r include=TRUE, tidy=FALSE,eval=TRUE}

# Example:
a = c(5,7,2,9)
ifelse(a %% 2 == 0,"even","odd")

#OR

x <- 5
if(x > 0){
  print("Positive number")
}
```

\vspace{12pt}

\textbf{\large apply} 

To do a task many times using vectorization. Much faster then a for loop.

```{r include=TRUE, tidy=FALSE,eval=TRUE}
x <- matrix(rnorm(400*4000), ncol=400)

# The 1 indicates to the function 'max' on the rows of the object 'x'. A 2 would 
# make this function should be applied to each column of the object 'x'.
mx2 <- apply(x, 1, max)

```

Another Example

```{r include=TRUE, tidy=FALSE,eval=TRUE}

dataTable <- cbind(x1 = 1, x2 = (c(4:0, 2:6)))
dataTable

# Sum each of the rows of the data table
apply(dataTable, MARGIN = 1, FUN = sum)  


# Now sum the columns
apply(dataTable, MARGIN = 2, FUN = sum)

```

\vspace{12pt}

\textbf{\large lapply}

To do a task many times using vectorization for each list element.

```{r include=TRUE, tidy=FALSE,eval=TRUE}
list1 <- list(1:20, 1:5, 1:100)
listOutput <- lapply(list1, FUN = quantile, probs = c(0.025, 0.975))
listOutput
is.list(listOutput) 
```


\textbf{\large sapply} 

takes lists, vectors or data frames as input and gives output in vector or matrix.

```{r include=TRUE, tidy=FALSE,eval=TRUE}
listOutput2 <- sapply(list1, FUN = quantile, probs = c(0.025, 0.975))
listOutput2 
is.list(listOutput2) 
is.matrix(listOutput2) 

```
\vspace{12pt}

\textbf{\large mapply}

allows you to apply a function to elements of a matrix.

```{r include=TRUE, tidy=FALSE,eval=TRUE}
l1 <- list(a = c(1:10), b = c(11:20))
l2 <- list(c = c(80:89), d = c(90:99))

# sum the corresponding elements of l1 and l2
mapply(sum, l1$a, l1$b, l2$c, l2$d)

```

\vspace{12pt}
