---
title:  " "
format:
  html: 
    theme: cosmo
    fontsize: 1.5em
    linestretch: 1.15
---


## FW 680A4 (Wildlife Ecology Modeling)

We will regularly use the [R programming language](https://www.r-project.org/about.html) to manipulate data and objects, create figures, and use functions.


<br>

I expect students to know some fundamentals of base R.


<br>

I am not expecting advanced knowledge of highly specialized and efficient programming.


<br>

More so, I don't want learning base R to be an impediment of understanding what we are trying to accomplish.


## Objects {.scrollable}

**Some things I expect students to know**

- what an **object ** is

- how to subset and manipulate different types of objects

  - vectors, lists, matrices, dataframes, etc.

#### E.g.....

::: {.fragment}

```{r}
#| echo: TRUE
#| eval: TRUE
#| tidy: true

# A vector of integers stored as an object...

vec = seq(1,10)

vec


# A Matrix...

mat = matrix(vec,nrow=2,ncol=5)

mat[2,]

# A list

my.list = list(vec, mat)

my.list[[2]]

```

:::

## Access {.scrollable}

Accessing a dataframe

#### E.g.....

```{r}
#| echo: TRUE
#| eval: TRUE
#| tidy: true

head(mtcars)
```

## Access {.scrollable}

Accessing a dataframe

#### E.g.....

```{r}
#| echo: TRUE
#| eval: TRUE
#| tidy: true

mtcars$mpg

mtcars[,1]

mtcars['mpg']
```


## Indexing {.scrollable}

Indexing a dataframe

#### E.g.....

```{r}
#| echo: TRUE
#| eval: TRUE
#| tidy: true

# find the location of certain values

which(mtcars$mpg>20)

# manipulating those elements

mtcars$mpg[which(mtcars$mpg>20)] = NA
mtcars$mpg

```

## function {.scrollable}

What a **function ** is

  - how to specify attributes of a function
  - how to wrap functions

#### E.g.....

::: {.fragment}

```{r}
#| echo: TRUE
#| eval: TRUE
#| tidy: true

# Specifying attributes of a function

mean(mtcars$mpg,
     na.rm=TRUE
     )
```

:::

## function {.scrollable}

#### E.g.....

```{r}
#| echo: TRUE
#| eval: TRUE
#| tidy: true

# Wrapping functions

sim = rnorm(10, 
            mean = mean(mtcars$mpg,na.rm=TRUE),
            sd = sd(mtcars$mpg,na.rm=TRUE)
)
 
sim
```


## Last

Everyone codes differently.

<br>

You may use **tidy** over base R. That is fine. 

<br>

If you are concerned about coding, talk to Brian (202A Wagar)

<br>

If you need resources for learning R, please go [here](/FW680A4/) and click on the tab *Learning R*. 
