---
title: "Some Markdown and R Stuff"
author: "Brian Gerber"
date: "8/24/2022"
output: html_document
---

<style type="text/css">
body, td {
   font-size: 18px;
}
code.r{
  font-size: 20px;
}
pre {
  font-size: 20px
}
</style>

```{r setup, include=FALSE}
#setwd("~/Markdown")
setwd("C:/Users/bgerber/Google Drive/NRS520/Week 0")
knitr::opts_chunk$set(echo = TRUE)
```

# R Markdown

RStudio can help you use Markdown to create HTML/PDF/Word documents that can embed R code, HTML, CSS, latex (type setting language), and more. This is a good [website](https://www.markdownguide.org/) to checkout for markdown coding.

We will use markdown as a way accomplish homework. I would suggest to not try and get too fancy at first. Keep it simple. Write-text in RStudio as you would in any word processor. Write R code in R chunks. Use "echo=FALSE" and "echo=TRUE" for a code chink to either not print the code or to print the code. 

For example, let's create an R chunk and plot some data by simulating from a random variable `y`. Let's define this variable such that it can take on values of any real number from $-\infty$ to $\infty$. So, $y \in (-\infty, \infty)$ or $y \in \mathbb{R}$. 

More specifically, let's define `y` as having a probability density function of a Normal/Gaussian distribution, 

$$y \sim \text{Normal}(\mu, \sigma).$$

And stated more generally as,

$$f(y;\mu, \sigma) = [y|\mu, \theta].$$


```{r, chunk1}
#Simulate data from a standard normal distribution (mu = 0, sd = 1)
  y=rnorm(100,0,1)

#Create a histogram
  hist(y)
```

Let's create a new plot and not output the code. This time we will see the true probability distribution, not simulated values.

```{r, chunk2, echo=FALSE,fig.width=5,fig.align='center'}
#Plot the normal distrbution
  curve(dnorm(x,0,1),xlim=c(-5,5))
```

HTML code can be embedded, such as turning text a different <span style="color:red">color</span>.

Simple text formatting can be done directly, such as *italics* or **bolding**.

<hr style="border:2px solid gray">

# Equations 

Markdown uses latex to embed equation. For example, the expectation of random variable `y` is the integral from limits a to b of `y` multiplied by the probability density of `y`:

$$E[y] = \int_a^b y\times f(y; \mu, \sigma) dy$$
More so, you can use a lot of functionality of latex and make the expectation more clear as,
$$
\begin{align*} 
E[y] =& \int_{-\infty}^{\infty}  y\times \frac{1}{\sigma\sqrt(2\pi)}e^{-1/2(\frac{y-\mu}{\sigma})^2} dx \\
E[y] =& \mu.
\end{align*}
$$

<hr style="border:2px solid gray">

## Approximating an Integral

Let's understand this math a bit more. What we are doing is for every value of `y` possible ($-\infty$ to $\infty$), we multiple it by its probability density. Then we take all these values and add them (i.e., the integral). 

It's a bit easier to understand what the math is doing by making an approximation. 

We can approximate the integral by breaking it into pieces to calculate `y` times the $f(y; \mu, \sigma)$ and then sum all these values.

```{r, chunk3, echo=TRUE}
#Create a histogram and save the values
  set.seed(4341)
  y.hist=hist(rnorm(1000,0,1),breaks=8, freq=FALSE,ylim=c(0,0.4))

#Plot the true probabiltiy denisty.
  curve(dnorm(x,0,1),add=TRUE,lwd=3,col=2)

#Get the y-axis values (probability density) for each value of the mid-point of each bin. 
  mid.point.x=y.hist$mids
  mid.point.x

  prob.y=y.hist$density
  prob.y

#multiple these value and sum them
  sum(prob.y*mid.point.x)
```

### Does this answer make sense?

What if you did the same approximation but with a lot more simulated values and bins.


<hr style="border:2px solid gray">

# An Assignment

Please create a markdown document (HTML/PDF) that implements the R code outlined below along with text explanation of what the code is doing. 

We will start this process as a group in R and you will document and explain the code by creating a markdown file. <span><u>Use Latex to create at least one equation of your choosing.</u></span>

## <span style="color:blue">What to include</span>

### <span style="color:red">Create Data</span>

- Simulate values of random variable `y` from your favorite probability distribution (e.g., `rnorm`, `rgamma`,`rbeta`,`rexp``), creating a matrix that is 1000 rows by 200 columns.

### <span style="color:red">apply function</span>

- use the `apply` function to calculate the `range` (this is the function name) of values for each row.

- next, use the `apply` function to calculate the standard deviation (`sd`) of all the rows of `y`.


### <span style="color:red">Your own function</span>

- create a new function `(myfunc=function(x){does stuff to x})` that when provided a vector (x), subtracts the max value from the min value of x. Use your new function in the `apply` function to calculate the difference b/w the max and min values for all the rows of `y`.

- make a histogram of your new values.

### <span style="color:red">For loops</span>

- create a for loop, looping over `y` from row 1 to `nrow(y)`. Within the loop, for each ith row, calculate the mean and save this value in a new vector. Plot all the row means in a histogram.

- next, find all the row means of `y` using the `apply` function. Plot these results with a histogram. Is it the same plot as the previous one?


### <span style="color:red">Mean, SD, mapply, lapply</span>

- Create a vector (called mu) that is a sequence of values representing different means. Use the seq function. Lower value should be zero, upper value should be 1000.

- Create a vector (called sigma) that has two values, representing different standard deviations. Do not use zero as one of the numbers.

- Use the `expand.grid` function to find all possible combinations of mu and sigma.

- create a new function that has two inputs (`a` and `b`). The function should draw 1000 random values from the Normal distribution (`rnorm`) for a given mean, `a`, and stdev `b`. The function should return the values in a list.

- Use the`mapply` function to input your combinations of mu/stdev and execute your new function. Your new object should contain many lists of 1000 random values for each combination of mu and sigma values.  

- Use the `lapply` function to apply a new function you create that estimates the mean and stdev of each list of 1000 values from your object.

- compare your estimated means and stdevs with your inputted ones (mu, sigma). Are you surprised by the difference or not?



